Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts

### 2017-12-03 article/article.tex
adiantada a sintese de textos

Possible syntactic structures are unlimited.
For a noun or verb phrase, one might quantify the possibilities
by considering all words but the particles that are nothing else
(i.e. li, e, la, pi, a, o, anu, en, seme, mu).
This yields $120-10=110 words$.
As discussed in Section~\ref{syntax},
I advocate that, although such words are classified in the dictionary,
they might be used indistinctly as nouns, adjectives,
verbs and adverbs.
Thus, one have 110 possibilities of noun and verb phrases with
one word, $110^2=12100$ possibilities with two words,
$110^3=1331000$ with three words and so on.
Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
(verb), object (noun) and prepositional (noun or verb) phrases of a sentence,
and assume that the sentence has at most one prepositional phrase,
one might quantify the possibilities using the formula:
$\delta = 110^n\times 110^v\times 10^o \times 5\times 110^p$,
where $5$ stands for the possible prepositions.
To account for the particles, one possibility is
to assume for simplicity that one might use none or one particle
(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
at each phrase, yielding:
$\delta = 110^n\times 110^v\times 110^o \times 5\times 110^p\times 9^4$.
For example, assume $n=v=p=o=1$, then
$\delta=110^1\times 110^1\times 110^1 \times 5\times 110^1\times 9^4=
4.80298^{12}$
i.e. almost 5 trillion possible
sentences with single word phrases (i.e. no adjectives or adverbs),
while allowing only one particle per phrase
and only one predicate phrase 
(e.g. 'ona li moku e soweli lon supa').
The general case is implemented as a function
in the script \tttt{makeStatistics.py}~\cite{tokipona}

>>>>

Possible syntactic structures are unlimited.
For a noun or verb phrase, one might quantify the possibilities
by considering all words but the particles that are nothing else
(i.e. li, e, la, pi, a, o, anu, en, seme, mu)
and the presotions that are nothing else (i.e. kepeken, lon, tan).
This yields $120-13=107 words$.
As discussed in Section~\ref{syntax},
I advocate that, although such words are classified in the dictionary,
they might be used indistinctly as nouns, adjectives,
verbs and adverbs.
Thus, one have 107 possibilities of noun and verb phrases with
one word, $107^2=11449$ possibilities with two words,
$107^3=1225043$ with three words and so on.
Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
(verb), object (noun) and prepositional (noun or verb) phrases of a sentence,
and assume that the sentence has at most one prepositional phrase,
one might quantify the possibilities using the formula:
$\delta = 107^n\times 107^v\times 10^o \times 5\times 107^p$,
where $5$ stands for the possible prepositions.
To account for the particles, one possibility is
to assume for simplicity that one might use none or one particle
(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
at each phrase, yielding:
$\delta = 107^n\times 107^v\times 107^o \times 5\times 107^p\times 9^4$.
For example, assume $n=v=o=p=1$, then
$\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
 4.300066\times 10^+12$
i.e. more then 4 trillion possible
sentences with single word phrases (i.e. no adjectives or adverbs),
while allowing only one particle per phrase
and only one predicate phrase 
(e.g. 'ona li moku e soweli lon supa').
The general case is implemented as a function
in the script \tttt{makeStatistics.py}~\cite{tokipona}
